Type theories as quotient inductive-recursive types

### 2025-02-20 tex/main.tex
Polish and fix the notation for substitution up to Section A

\supplement{Formal Agda development: \url{https://github.com/L-TChen/TTasQIIRT}./}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
%\supplementdetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g. Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional
\acknowledgements{We'd like to thank Hsiang-Shang Ko and Meven Lennon-Bertrand.}%optional

@@

Metaprogramming and mechanised metatheory in a type-theoretic proof assistant relies on having an internal representation of type theory within type theory itself.
%
In the spirit of type theory, we would like to represent well scoped and well typed terms only --- this also has technical advantages, such as theorem statements not having to deal with impossible cases, or separately having to prove wellformedness lemmas.
%
Doing this effectively seems to require sophisticated features of the host type theory, most notably advanced inductive definitions.
%
For example, inductive families~\cite{Dybjer1994} can be used to represent the simply typed $\lambda$-calculus~\cite{Altenkirch1999}, while some kind of simultaneous definitions such as inductive-recursive~\cite{Dybjer2000} or inductive-inductive~\cite{Nordvall-Forsberg2014} definitions seem to be required for dependent type theory~\cite{Danielsson2006,Chapman2009}.
%
This is not just a matter of pure strength, but also expressivity and convenience: while it is known how to reduce inductive-inductive definitions to inductive families in principle~\cite{Kaposi2020a}, the resulting development would be considerably harder to work with, most likely exhausting the resources of both the proof assistant and its user.

@@

A typical workaround in practice \cite{Kaposi2019,Kaposi2024a} is to add equality constructors as rewrite rules after an inductive definition and further identify equality constructors with the reflxivity proof.
%
As the host theory is assumed to satisfy the uniqueness of identity proof and function extensionality, this workaround can be justified by conservativity of equality reflection over intensional type theory~\cite{Hofmann1996,Winterhalter2019,Kapulkin2025}.
%
However, this workaround is computationally ill-behaved, as confluence does not hold in general.
Also, the presentation often differs from the actual definition of type theory in intensional type theory.
%
Any method that reduces the friction introduced by the use of equality constructors without compromising on computational properties and the gap between formalisation and presentation seem worth exploring.
%

@@

Quotient inductive types (QITs) are a special case of higher inductive types~\cite{Lumsdaine2020}, where constructors can introduce not only new elements but also new equalities between the introduced elements.
%
%Since we assume the uniqueness of identity proofs, all QITs are implicitly truncated to be sets in the sense of homotopy type theory.
%
For example, the type $\JList{A}$ of \emph{join lists} can be given by the QIT in \cref{fig:QIT-join-list}.
\begin{figure}
\begin{minipage}{.35\textwidth}
\begin{alignat*}{3}
  \Jnil                    & : \implicit{A}      && \JList{A} \\
  \Jsing{\blank}           & : \implicit{A}      && A \to \JList{A} \\
  \Jconcat{\blank}{\blank} & : \implicit{A}      && \JList{A} \to \JList{A} \to \JList{A}
\end{alignat*}
\end{minipage}
\begin{minipage}{.64\textwidth}
\begin{alignat*}{3}
  \mathsf{idr}             & : \implicit{\ell}\; && (\Jconcat{\ell}{\Jnil}) =^{\JList{A}} \ell \\
  \mathsf{idl}             & : \implicit{\ell}\; && (\Jconcat{\Jnil}{\ell}) =^{\JList{A}} \ell \\
  \mathsf{assoc}           & : \implicit{\ell_0, \ell_1, \ell_2}\; && \Jconcat{(\Jconcat{\ell_0}{\ell_1})}{\ell_2} =^{\JList{A}} \Jconcat{\ell_0}{(\Jconcat{\ell_1}{\ell_2})}
\end{alignat*}
\end{minipage}
\caption{The quotient inductive type of join lists}
\label{fig:QIT-join-list}
\end{figure}
%\vskip\belowdisplayskip\noindent
A quotient inductive-inductive type~\cite{Altenkirch2018} (QIIT) further allows the simultaneous definition of a type $A$ and a type family $B : A \to \Set$ defined inductively.
% --- this is different from an inductive-recursive definition of $A$ and $F : A \to D$, where $F$ is defined recursively.
%; indeed, inductive-inductive definitions only make sense when $D = \Set$.
The metatheory of QIITs has been studied by Altenkirch et al.~\cite{Kaposi2018,Kaposi2019,Kovacs2020}.

@@

\subsection{Definitions by overlapping patterns} \label{sec:meta:overlapping}
In a proof assistant such as \Agda, functions are defined using pattern matching and elaborated into case trees, which correspond to the eliminator, following a \emph{first-match} semantics~\cite{Cockx2020a} and employing termination checking.
%This elaboration offers conventional notation, while making computational behaviour opaque.
% and favouring asymmetric definitions.
For example, the two definitions of addition $\blank  + \blank$, $\blank  +' \blank $
%
%\begin{figure}[b]%
\begin{minipage}{.48\textwidth}
  \begin{alignat*}{5}
%    \blank  & + \blank     && : \N \to \N \to \N \\
    0       &  + m         && = m  \\
  n         &  + 0         && = m  \\
  (\suc\;n) &  + m         && = \suc\;(n + m) \\
  n         &  + (\suc\;m) && = \suc\;(n + m)
\end{alignat*}
\end{minipage}
\begin{minipage}{.5\textwidth}
\begin{alignat*}{5}
%  \blank         & +'\blank     && : \N \to \N \to \N \\
  0              & +' m         && \defeq m  \\
  (\suc\;n)      & +' 0         && \defeq \suc\;n \\
  (\suc\;n)      & +' (\suc\;m) && \defeq \suc\;(n +' \suc\;m)
\end{alignat*}
\end{minipage}
\vskip\belowdisplayskip\noindent
%\caption{Two seemingly different definitions of addition}
%\label{fig:additions}
%\end{figure}
are the same after elaboration in \Agda: eliminating the first argument of $+$ distinguishes $0$ and $\suc\;n$ and refines the second and the third clauses into clauses of $+'$, making the fourth clause useless.
Each clause of $+'$ is strict while clauses of $+$ are not necessarily strict.
%This semantics for function definitions decides how definitions reduce and therefore how identities are proved.

@@

A substitution $\sigma : \Sub\;\Gamma\;\Delta$ acts on types $A : \Ty\;\Delta$ over $\Delta$ to ``pull them back'' to types $\sub{A}{\sigma} : \Ty\;\Gamma$ over $\Gamma$. That is, the action of type substitution can be seen as a constructor
\begin{alignat*}{3}
  \sub{\blank}{\blank} &: \implicit{\Gamma, \Delta} \; \Ty\;\Delta \to \Sub\;\Gamma\;\Delta \to \Ty\;\Gamma .
\end{alignat*}
%
Since $\Ty$ is inductively defined, we add a base case $\UU : \implicit{\Gamma} \Ty\,\Gamma$ to ensure that $\Ty\;\Gamma$ is not empty.
%for every $\Gamma$:
%
%If we were defining the notion of cwfs, type substitution would be enough for $\Ty$.
%For our inductive definition (i.e.\ the initial cwf), however, we need a base case in $\Ty\;\Gamma$, otherwise the type $\Ty$ would be empty as well as other types.
%Hence another constructor is introduced:
%\begin{alignat*}{3}
%  \UU & : \implicit{\Gamma} & \Ty\, \Gamma.
%\end{alignat*}
For now, $\UU$ serves as a constant, but we will reuse $\UU$ for the type of small types later.

Substitutions from $\Gamma$ to $\Delta = \emptyctx, A_1, \ldots, A_n$ can be understood intuitively as lists of terms $t_i : A_i$ in context $\Gamma$ for each type $A_i$ in $\Delta$: we have the empty substitution $\emptysub$, and substitution extensions $\sigma, t$ which substitutes the most recently bound variable by the term~$t$:
\[
  \emptysub : \implicit{\Gamma}\;\Sub\;\Gamma\;\emptyctx
  \qquad\qquad
  \blank,\blank : \implicit{\Gamma, \Delta, A}\; (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\;\Gamma\;(\sub{A}{\sigma}) \to \Sub\;\Gamma\;(\Delta, A).
\]
Note that the type substitution $\sub{A}{\sigma}$ is needed, because $A$ is well-formed in context $\Delta$ rather than in $\Gamma$.
We also ask for identity substitutions $\idS$ and compositions $\sigma;\tau$.
%because the substitution calculus is supposed to be the initial \emph{category} with families:
\[
  \idS : \implicit{\Gamma}\; \Sub\;\Gamma\;\Gamma
  \qquad\qquad \blank;\blank : \implicit{\Gamma, \Delta, \Theta}\;\Sub\;\Gamma\;\Delta \to \Sub\;\Delta\;\Theta \to \Sub\;\Gamma\;\Theta.
\]
These need to satisfy laws, which we will introduce later as equality constructors.
Context extension is characterised by its projections, which forgets and picks out the most recently substituted term respectively:
%Context comprehension is given by projections from \emph{non-empty} substitutions $\sigma : \Sub\;\Gamma\;(\Delta, A)$ for its \emph{tail} and its \emph{head}:
\begin{alignat*}{3}
  \pi_1 & : \implicit{\Gamma, \Delta, A}\;\Sub\;\Gamma\;(\Delta, A) \to \Sub\;\Gamma\;\Delta \\
  \pi_2 & : \implicit{\Gamma, \Delta, A}\;(\sigma : \Sub\;\Gamma\;(\Delta, A)) \to \Tm\;\Gamma\;(\sub{A}{\pi_1\,\sigma})
\end{alignat*}
%
Again, a type substitution $\sub{A}{\pi_1\,\sigma}$ is needed for~$\pi_2$, because $A$ is well-formed in context $\Delta$ instead of $\Gamma$.
Finally, we have term substitution, which given $\sigma : \Sub\;\Gamma\;\Delta$ turns terms in $\Delta$ to terms in $\Gamma$, appropriately preserving the type, which is also substituted:
\begin{alignat*}{3}
  \sub{\blank}{\blank} & : \implicit{\Gamma,\Delta, A}\;(\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Delta\;A \to \Tm\;\Gamma\;(\sub{A}{\sigma}).
\end{alignat*}
%Note that $\sub{\blank}{\blank}$ is overloaded for type and term substitution.

The structural rules for type substitution are stipulated by following equality constructors:
\begin{align*}
  [\idS]_{\Ty} &: \implicit{\Gamma, A} \; \sub{A}{\idS} =^{\Ty\,\Gamma} A
  \qquad\qquad
  [;]_{\Ty}  : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A}\; \sub{A}{\sigma ; \tau} =^{\Ty\,\Gamma} \sub{(\sub{A}{\tau})}{\sigma}
  \\
  []\UU &: \implicit{\Gamma, \Delta, \sigma} \; \sub{\UU}{\sigma} =^{\Ty\,\Gamma} \UU
\end{align*}
%
%\begin{alignat*}{3}
%  [\idS]_{\Ty} & : \implicit{\Gamma, A}                               && \sub{A}{\idS}           && =^{\Ty\,\Gamma}\;A \\
%  [;]_{\Ty}    & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A}\;&& \sub{A}{\sigma ; \tau}  && =^{\Ty\,\Gamma}\; \sub{(\sub{A}{\tau})}{\sigma} \\
%  []\UU      & : \implicit{\Gamma, \Delta, \sigma}                    && \sub{\UU}{\sigma}        && =^{\Ty\,\Gamma}\;\UU
%\end{alignat*}

@@

The use of transports in the formal definition fixes type mismatches but hinders equational reasoning about these terms, even for simple facts such as the one below.
\begin{example}\label{ex:pi2-comp}
  Given substitutions $\sigma : \Sub\;\Gamma\; \Delta$ and $\tau : \Sub\;\Delta\;(\Theta, A)$ for any $A : \Ty\,\Theta$, we may apply the projection $\pi_2$ to the composite $(\sigma; \tau)$ to access the first term $\pi_2(\sigma; \tau)$ of type $\sub{A}{\sigma;\tau}$ under the context $\Gamma$, and this term should be equal to the first term $\pi_2\,\tau$ of $\tau$ after applying the substitution $\sigma$. 
  In short, the following identity apparently holds
  \[
    \pi_2\,(\sigma ; \tau) = \sub{(\pi_2\,\tau)}{\sigma}
  \]
  and is checked by a back-of-the-envelope calculation
  \begin{equation} \label{eq:pi2-comp-proof}
    \pi_2\,(\sigma ; \tau) 
    = \pi_2\,(\sigma; (\pi_1\,\tau, \pi_2\,\tau))
    = \pi_2\,(\sigma;\pi_1\,\tau, \sub{(\pi_2\,\tau)}{\sigma})
    = \sub{(\pi_2\,\tau)}{\sigma}.
  \end{equation}
  Yet, the left-hand side is a term of type $\sub{A}{\pi_1\,(\sigma;\tau)}$, but the other is $\sub{\sub{A}{\pi_1\,\tau}}{\sigma}$.
  Hence the above identity does not even make sense, since their types do not match.
  Alas, instead, we have to write $\pi_2\,(\sigma ; \tau) =^{\Tm\,\Gamma}_{p} \sub{(\pi_2\,\tau)}{\sigma}$ for some $p: \sub{A}{\pi_1(\sigma; \tau)} = \sub{\sub{A}{\pi_1\tau}}{\sigma}$ or, equivalently
  \begin{equation}\label{eq:pi2-comp-real-proof}
    (\sub{A}{\pi_1(\sigma; \tau)}, \pi_2(\sigma; \tau)) =^{(A : \Ty\,\Gamma) \times (\Tm \Gamma A)} (\sub{\sub{A}{\pi_1\,\tau}}{\sigma}, \sub{(\pi_2\,\tau)}{\sigma})
  \end{equation}
  as inhabitants of a $\Sigma$-type, so we can reason about term equalities along with type equalities.
  Moreover, in~\eqref{eq:pi2-comp-proof} we have used the rule $\mathsf{concat}$ which introduced another transported term, so we will have to remove that $\transp$ to derive the right hand side.

@@

\subsection{Adding a universe} \label{subsec:SC+U}
Now we extend substitution calculus with an empty universe $\UU$ of small types~$\El\,u$.
In a QIIT definition \cite{Altenkirch2016a}, this extension adds the following constructors
\begin{alignat*}{3}
  \El   & : \implicit{\Gamma} && \Tm\,\Gamma\;\UU \to \Ty\,\Gamma \\
  []\El & : \implicit{\Gamma, \Delta, \sigma, u} && \sub{(\El\, u)}{\sigma} =^{\Tm\;\Gamma\;\UU} \El\,(\alert{\transfib{\Tm\;\Gamma}{[]\UU}{\color{black}\sub{u}{\sigma}}})
\end{alignat*}
%as the type $\UU$ of universe and its structural rule $[]\UU$ have been added to avoid a trivial inductive type.
The type substitution $\sub{(\El\,u)}{\sigma}$ is, informally, the type of the term substitution $\sub{u}{\sigma}$ for $u : \Tm\;\Delta\;\UU$, and the transport is needed to coerce $\sub{u}{\sigma} : \Tm\;\Gamma\;(\sub{\UU}{\sigma})$ into $\Tm\;\Gamma\;\UU$.
%
In our QIIR definition, $\transfib{\Tm\;\Gamma}{[]\UU}{\blank}$ is no longer needed, since $[]\UU$ becomes strict. 
Likewise, the equality constructor $[]\El$ may be replaced by the reduction
\begin{alignat*}{3}
  \subTy{(\El\, u)}{\sigma} & \reduce \El\,(\sub{u}{\sigma})
\end{alignat*}
Yet, this naive change breaks local confluence!
The problem is that there is now additional interaction between types and terms.
% Observe that
The local peak
\(
  \El\,u \Leftarrow \subTy{(\El\,u)}{\idS} \reduce \El(\sub{u}{\idS})
\)
cannot reduce to the same term, since the term substitution $\sub{u}{\idS}$ is merely a constructor.
%
To repair local confluence, we add a simultaneously defined term substitution function $\subTm{\blank}{\blank} : \implicit{\Gamma, \Delta, A}\;\Tm\,\Delta\,A \to (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Gamma\,(\subTy{A}{\sigma})$, different from the constructor $\sub{\blank}{\blank}$, with the following reductions:
\begin{alignat*}{3}
%  \subTm{\blank}{\blank} & : \implicit{\Gamma, \Delta, A}\;\Tm\,\Delta\,A \to (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Gamma\,(\subTy{A}{\sigma}) \\
\subTm{t}{\idS}          & \reduce t \\
\subTm{t}{\sigma ; \tau} & \reduce \subTm{(\subTm{t}{\tau})}{\sigma} \\
\subTm{t}{\pi_1(\sigma, s)} & \reduce \subTm{t}{\sigma} \\
\subTm{t}{\pi_1(\sigma; \tau)} & \reduce \subTm{(\subTm{t}{\pi_1\tau})}{\sigma} \\
\subTm{t}{\sigma}        & \reduce \sub{t}{\sigma}, \quad \text{otherwise}
\end{alignat*}
This reduces substitutions that occur in \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4} recursively so that the reduction can be propagated from types to terms.
The remaining ``default'' cases reduce to the constructor $\sub{t}{\sigma}$.
Note that $\subTm{\blank}{\blank}$ can be defined with the standard first-match pattern matching semantics, since there are no overlapping patterns.
\begin{remark}
  The function $\subTm{\blank}{\blank}$ can be understood as a \emph{smart constructor} for the explicit term substitution $\sub{\blank}{\blank}$, a well-known idiom in functional programming: the function $\subTm{\blank}{\blank}$ computes certain cases away, but mostly reduces to the non-smart constructor $\sub{\blank}{\blank}$.
\end{remark}

>>>>

\supplement{Formal Agda development: \url{https://github.com/L-TChen/TTasQIIRT}./}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
%\supplementdetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g. Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional
\acknowledgements{We'd like to thank Hsiang-Shang Ko for useful discussion.}%optional

@@

Metaprogramming and mechanised metatheory in a type-theoretic proof assistant relies on having an internal representation of type theory within type theory itself.
%
In the spirit of type theory, we would like to represent well typed terms only --- this has technical advantages, such as theorem statements not having to deal with impossible cases, or separately having to prove wellformedness lemmas.
%
Doing this effectively seems to require sophisticated features of the host type theory, most notably advanced inductive definitions.
%
For example, inductive families~\cite{Dybjer1994} can be used to represent the simply typed $\lambda$-calculus~\cite{Altenkirch1999}, while some kind of simultaneous definitions such as inductive-recursive~\cite{Dybjer2000} or inductive-inductive~\cite{Nordvall-Forsberg2014} definitions are used for dependent type theory~\cite{Danielsson2006,Chapman2009}.
%
This is not just a matter of pure strength, but also expressivity and convenience: while it is known how to reduce inductive-inductive definitions to inductive families in principle~\cite{Kaposi2020a}, the resulting development would be considerably harder to work with, most likely exhausting the resources of both the proof assistant and its user.

@@

A typical workaround in practice \cite{Kaposi2019,Kaposi2024a} is to add equality constructors as rewrite rules~\cite{Cockx2021,Cockx2020} after an inductive definition.
%
As the host theory is assumed to satisfy the uniqueness of identity proof and function extensionality, this workaround can be justified by conservativity of equality reflection over intensional type theory~\cite{Hofmann1996,Winterhalter2019,Kapulkin2025}.
%
However, this workaround is computationally ill-behaved, as confluence does not hold in general.
Also, the presentation often differs from the actual definition in type theory.
%
Any method that reduces the friction introduced by equality constructors without compromising on computational properties and the gap between formalisation and presentation seem worth exploring.
%

@@

Quotient inductive types (QITs) are a special case of higher inductive types~\cite{Lumsdaine2020}, where constructors can introduce not only new elements but also new equalities between the introduced elements.
%
%Since we assume the uniqueness of identity proofs, all QITs are implicitly truncated to be sets in the sense of homotopy type theory.
%
For example, the type $\JList{A}$ of \emph{join lists} can be given by the QIT in \cref{fig:QIT-join-list}.
\begin{figure}[b]
\begin{minipage}{.35\textwidth}
\begin{alignat*}{3}
  \Jnil                    & : \implicit{A}\;      && \JList{A} \\
  \Jsing{\blank}           & : \implicit{A}\;      && A \to \JList{A} \\
  \Jconcat{\blank}{\blank} & : \implicit{A}\;      && \JList{A} \to \JList{A} \to \JList{A}
\end{alignat*}
\end{minipage}
\begin{minipage}{.64\textwidth}
\begin{alignat*}{3}
  \mathsf{idr}             & : \implicit{\ell}\; && (\Jconcat{\ell}{\Jnil}) =^{\JList{A}} \ell \\
  \mathsf{idl}             & : \implicit{\ell}\; && (\Jconcat{\Jnil}{\ell}) =^{\JList{A}} \ell \\
  \mathsf{assoc}           & : \implicit{\ell_0, \ell_1, \ell_2}\; && \Jconcat{(\Jconcat{\ell_0}{\ell_1})}{\ell_2} =^{\JList{A}} \Jconcat{\ell_0}{(\Jconcat{\ell_1}{\ell_2})}
\end{alignat*}
\end{minipage}
\caption{The quotient inductive type of join lists}
\label{fig:QIT-join-list}
\end{figure}
%\vskip\belowdisplayskip\noindent
A quotient inductive-inductive type~\cite{Altenkirch2018} (QIIT) further allows the simultaneous definition of a type $A$ and a type family $B : A \to \Set$ defined inductively.
% --- this is different from an inductive-recursive definition of $A$ and $F : A \to D$, where $F$ is defined recursively.
%; indeed, inductive-inductive definitions only make sense when $D = \Set$.
The metatheory of QIITs has been studied by Altenkirch et al.~\cite{Kaposi2018,Kaposi2019,Kovacs2020}.

@@

\subsection{Definitions by overlapping patterns} \label{sec:meta:overlapping}
In a proof assistant such as \Agda, functions are defined using pattern matching and elaborated into case trees, which correspond to the eliminator, following a \emph{first-match} semantics~\cite{Cockx2020a} and employing termination checking.
%This elaboration offers conventional notation, while making computational behaviour opaque.
% and favouring asymmetric definitions.
For example, the two definitions of addition $+$ and $+'$
%
%\begin{figure}[b]%
\begin{minipage}{.48\textwidth}
  \begin{alignat*}{5}
%    \blank  & + \blank     && : \N \to \N \to \N \\
    0       &  + m         && = m  \\
  n         &  + 0         && = m  \\
  (\suc\;n) &  + m         && = \suc\;(n + m) \\
  n         &  + (\suc\;m) && = \suc\;(n + m)
\end{alignat*}
\end{minipage}
\begin{minipage}{.5\textwidth}
\begin{alignat*}{5}
%  \blank         & +'\blank     && : \N \to \N \to \N \\
  0              & +' m         && \defeq m  \\
  (\suc\;n)      & +' 0         && \defeq \suc\;n \\
  (\suc\;n)      & +' (\suc\;m) && \defeq \suc\;(n +' \suc\;m)
\end{alignat*}
\end{minipage}
\vskip\belowdisplayskip\noindent
%\caption{Two seemingly different definitions of addition}
%\label{fig:additions}
%\end{figure}
are the same after elaboration in \Agda: eliminating the first argument of $+$ distinguishes $0$ and $\suc\;n$ and refines the second and the third clauses into clauses of $+'$, making the fourth clause useless.
Each clause of $+'$ is strict while clauses of $+$ are not necessarily strict.
%This semantics for function definitions decides how definitions reduce and therefore how identities are proved.

@@

A substitution $\sigma : \Sub\;\Gamma\;\Delta$ acts on types $A : \Ty\;\Delta$ over $\Delta$ to `pull them back' to types $\sub{A}{\sigma} : \Ty\;\Gamma$ over $\Gamma$. That is, the action of type substitution can be seen as a constructor
\begin{alignat*}{3}
  \sub{\blank}{\blank} &: \implicit{\Gamma, \Delta} \; \Ty\;\Delta \to \Sub\;\Gamma\;\Delta \to \Ty\;\Gamma .
\end{alignat*}
%
Since $\Ty$ is inductively defined, we add a base case $\UU : \implicit{\Gamma} \Ty\,\Gamma$ to ensure that $\Ty\;\Gamma$ is not empty.
%for every $\Gamma$:
%
%If we were defining the notion of cwfs, type substitution would be enough for $\Ty$.
%For our inductive definition (i.e.\ the initial cwf), however, we need a base case in $\Ty\;\Gamma$, otherwise the type $\Ty$ would be empty as well as other types.
%Hence another constructor is introduced:
%\begin{alignat*}{3}
%  \UU & : \implicit{\Gamma} & \Ty\, \Gamma.
%\end{alignat*}
For now, $\UU$ serves as a constant, but we will reuse $\UU$ for the type of small types later.

Substitutions from $\Gamma$ to $\Delta = \emptyctx, A_1, \ldots, A_n$ can be understood intuitively as lists of terms $t_i : A_i$ in context $\Gamma$ for each type $A_i$ in $\Delta$: we have the empty substitution $\emptysub$, and substitution extensions $\sigma, t$ which substitutes the most recently bound variable by the term~$t$:
\[
  \emptysub : \implicit{\Gamma}\;\Sub\;\Gamma\;\emptyctx
  \qquad\qquad
  \blank,\blank : \implicit{\Gamma, \Delta, A}\; (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\;\Gamma\;(\sub{A}{\sigma}) \to \Sub\;\Gamma\;(\Delta, A).
\]
Note that the type substitution $\sub{A}{\sigma}$ is needed, because $A$ is well-formed in context $\Delta$ rather than in $\Gamma$.
We also ask for identity substitutions $\idS$ and compositions $\sigma;\tau$.
%because the substitution calculus is supposed to be the initial \emph{category} with families:
\[
  \idS : \implicit{\Gamma}\; \Sub\;\Gamma\;\Gamma
  \qquad\qquad \blank;\blank : \implicit{\Gamma, \Delta, \Theta}\;\Sub\;\Gamma\;\Delta \to \Sub\;\Delta\;\Theta \to \Sub\;\Gamma\;\Theta.
\]
These need to satisfy laws, which we will introduce later as equality constructors.
Context extension is characterised by its projections, which forgets and picks out the most recently substituted term respectively:
%Context comprehension is given by projections from \emph{non-empty} substitutions $\sigma : \Sub\;\Gamma\;(\Delta, A)$ for its \emph{tail} and its \emph{head}:
\begin{alignat*}{3}
  \pi_1 & : \implicit{\Gamma, \Delta, A}\;\Sub\;\Gamma\;(\Delta, A) \to \Sub\;\Gamma\;\Delta \\
  \pi_2 & : \implicit{\Gamma, \Delta, A}\;(\sigma : \Sub\;\Gamma\;(\Delta, A)) \to \Tm\;\Gamma\;(\sub{A}{\pi_1\,\sigma})
\end{alignat*}
%
Again, a type substitution $\sub{A}{\pi_1\,\sigma}$ is needed for~$\pi_2$, because $A$ is well-formed in context $\Delta$ instead of $\Gamma$.
Finally, we have term substitution, which given $\sigma : \Sub\;\Gamma\;\Delta$ turns terms in $\Delta$ to terms in $\Gamma$, appropriately preserving the type, which is also substituted:
\begin{alignat*}{3}
  \sub{\blank}{\blank} & : \implicit{\Gamma,\Delta, A}\;\Tm\;\Delta\;A \to (\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Gamma\;(\sub{A}{\sigma}).
\end{alignat*}
%Note that $\sub{\blank}{\blank}$ is overloaded for type and term substitution.

The structural rules for type substitution are stipulated by following equality constructors:
\begin{alignat*}{9}
  [\idS]_{\Ty} &: \implicit{\Gamma, A} \; && \sub{A}{\idS} && =^{\Ty\,\Gamma} A
  \qquad\qquad
  [;]_{\Ty}  : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A}\; \sub{A}{\sigma ; \tau} =^{\Ty\,\Gamma} \sub{(\sub{A}{\tau})}{\sigma}
  \\
  []\UU &: \implicit{\Gamma, \Delta, \sigma} \; && \sub{\UU}{\sigma} && =^{\Ty\,\Gamma} \UU
\end{alignat*}
%
%\begin{alignat*}{3}
%  [\idS]_{\Ty} & : \implicit{\Gamma, A}                               && \sub{A}{\idS}           && =^{\Ty\,\Gamma}\;A \\
%  [;]_{\Ty}    & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A}\;&& \sub{A}{\sigma ; \tau}  && =^{\Ty\,\Gamma}\; \sub{(\sub{A}{\tau})}{\sigma} \\
%  []\UU      & : \implicit{\Gamma, \Delta, \sigma}                    && \sub{\UU}{\sigma}        && =^{\Ty\,\Gamma}\;\UU
%\end{alignat*}

@@

The use of transports in the formal definition fixes type mismatches but hinders equational reasoning about these terms, even for simple facts such as the one below.
\begin{example}\label{ex:pi2-comp}
  Given any substitution $\sigma : \Sub\;\Gamma\; \Delta$ and a non-empty substitution $\tau : \Sub\;\Delta\;(\Theta, A)$ for any $A : \Ty\,\Theta$, the first term of the composite $(\sigma; \tau)$ should intuitively be equal to the first term of $\tau$ after applying the substitution $\sigma$, i.e.\ the identity
  \[
    \pi_2\,(\sigma ; \tau) = \sub{(\pi_2\,\tau)}{\sigma}
  \]
  apparently holds and can checked by a back-of-the-envelope calculation
  \begin{equation} \label{eq:pi2-comp-proof}
    \pi_2\,(\sigma ; \tau) 
    = \pi_2\,(\sigma; (\pi_1\,\tau, \pi_2\,\tau))
    = \pi_2\,(\sigma;\pi_1\,\tau, \sub{(\pi_2\,\tau)}{\sigma})
    = \sub{(\pi_2\,\tau)}{\sigma}.
  \end{equation}
  Yet, the left-hand side is a term of type $\sub{A}{\pi_1\,(\sigma;\tau)}$, but the other is $\sub{\sub{A}{\pi_1\,\tau}}{\sigma}$.
  Hence the above identity does not even make sense, since their types do not match.
  Alas, instead, we have to write $\pi_2\,(\sigma ; \tau) =^{\Tm\,\Gamma}_{p} \sub{(\pi_2\,\tau)}{\sigma}$ for some $p: \sub{A}{\pi_1(\sigma; \tau)} = \sub{\sub{A}{\pi_1\tau}}{\sigma}$ or, equivalently
  \begin{equation}\label{eq:pi2-comp-real-proof}
    (\sub{A}{\pi_1(\sigma; \tau)}, \pi_2(\sigma; \tau)) =^{(A : \Ty\,\Gamma) \times (\Tm \Gamma A)} (\sub{\sub{A}{\pi_1\,\tau}}{\sigma}, \sub{(\pi_2\,\tau)}{\sigma})
  \end{equation}
  as inhabitants of a $\Sigma$-type, so we can reason about term equalities along with type equalities.
  Moreover, in~\eqref{eq:pi2-comp-proof} we have used the rule $\mathsf{concat}$ which introduced another transported term, so we will have to remove that $\transp$ to derive the right hand side.

@@

\subsection{Adding a universe} \label{subsec:SC+U}
Now we extend substitution calculus with an empty universe $\UU$ of small types~$\El\,u$.
In a QIIT definition \cite{Altenkirch2016a}, this extension adds the following constructors
\begin{alignat*}{3}
  \El   & : \implicit{\Gamma} && \Tm\,\Gamma\;\UU \to \Ty\,\Gamma \\
  []\El & : \implicit{\Gamma, \Delta, \sigma, u} && \sub{(\El\, u)}{\sigma} =^{\Tm\;\Gamma\;\UU} \El\,(\alert{\transfib{\Tm\;\Gamma}{[]\UU}{\color{black}\sub{u}{\sigma}}})
\end{alignat*}
%as the type $\UU$ of universe and its structural rule $[]\UU$ have been added to avoid a trivial inductive type.
The type substitution $\sub{(\El\,u)}{\sigma}$ is, informally, the type of the term substitution $\sub{u}{\sigma}$ for $u : \Tm\;\Delta\;\UU$, and the transport is needed to coerce $\sub{u}{\sigma} : \Tm\;\Gamma\;(\sub{\UU}{\sigma})$ into $\Tm\;\Gamma\;\UU$.
%
In our QIIR definition, $\transfib{\Tm\;\Gamma}{[]\UU}{\blank}$ is no longer needed, since $[]\UU$ becomes strict. 
Likewise, the equality constructor $[]\El$ may be replaced by the reduction
\begin{alignat*}{3}
  \subTy{(\El\, u)}{\sigma} & \reduce \El\,(\sub{u}{\sigma})
\end{alignat*}
Yet, this naive change breaks local confluence!
The problem is that there is now additional interaction between types and terms.
% Observe that
The local peak
\(
  \El\,u \Leftarrow \subTy{(\El\,u)}{\idS} \reduce \El(\sub{u}{\idS})
\)
cannot reduce to the same term, since the term substitution $\sub{u}{\idS}$ is merely a constructor.
%

To repair local confluence, we add a simultaneously defined term substitution function $\subTm{\blank}{\blank} : \implicit{\Gamma, \Delta, A}\;\Tm\,\Delta\,A \to (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Gamma\,(\subTy{A}{\sigma})$, different from the constructor $\sub{\blank}{\blank}$, with the following reductions:
\begin{alignat*}{3}
%  \subTm{\blank}{\blank} & : \implicit{\Gamma, \Delta, A}\;\Tm\,\Delta\,A \to (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Gamma\,(\subTy{A}{\sigma}) \\
\subTm{t}{\idS}          & \reduce t \\
\subTm{t}{\sigma ; \tau} & \reduce \subTm{(\subTm{t}{\tau})}{\sigma} \\
\subTm{t}{\pi_1(\sigma, s)} & \reduce \subTm{t}{\sigma} \\
\subTm{t}{\pi_1(\sigma; \tau)} & \reduce \subTm{(\subTm{t}{\pi_1\tau})}{\sigma} \\
\subTm{t}{\sigma}        & \reduce \sub{t}{\sigma}, \quad \text{otherwise}
\end{alignat*}
This reduces substitutions that occur in \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4} recursively so that the reduction can be propagated from types to terms.
The remaining `default' cases reduce to the constructor $\sub{t}{\sigma}$.
Note that $\subTm{\blank}{\blank}$ can be defined with the standard first-match pattern matching semantics, since there are no overlapping patterns.
\begin{remark}
  The function $\subTm{\blank}{\blank}$ can be understood as a \emph{smart constructor} for the explicit term substitution $\sub{\blank}{\blank}$, a well-known idiom in functional programming: the function $\subTm{\blank}{\blank}$ computes certain cases away, but mostly reduces to the non-smart constructor $\sub{\blank}{\blank}$.
\end{remark}
